---
title: "Fallstudie"
date: "2023-11-28"
output: pdf_document
---

```{r setup, include=FALSE}
# Load the DADA2 library
library(dada2)

# Set the working directory to your project folder
setwd("~/fallstudie")

# Load the conda environment with DADA2 (replace 'your_environment' with the actual environment name)
system("conda activate dada2")

# Create a vector of file names for forward and reverse reads
fnFs <- list.files(path = "01_rawdata", pattern = "_R1_001.fastq", full.names = TRUE)
fnRs <- list.files(path = "01_rawdata", pattern = "_R2_001.fastq", full.names = TRUE)
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)

```

##Quality control

```{r pressure, echo=FALSE}
# plot the quality control
plotQualityProfile(fnFs[1:2])
plotQualityProfile(fnRs[1:2])
```

```{r pressure, echo=FALSE}
#get the filtered values
filtFs <- file.path(path="~/fallstudie/03A_filtFs", paste0(sample.names, "_F_filt.fastq"))
filtRs <- file.path(path="~/fallstudie/03B_filtRs", paste0(sample.names, "_R_filt.fastq"))

names(filtFs) <- sample.names
names(filtRs) <- sample.names


out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(250,200),
                     maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
                     compress=TRUE, multithread=FALSE) # On Windows set multithread=FALSE ,
head(out)

#checking quality after triming
fnFs2 <- list.files(path = "03A_filtFs", pattern = "_F_filt.fastq", full.names = TRUE)
fnRs2 <- list.files(path = "03B_filtRs", pattern = "_R_filt.fastq", full.names = TRUE)
# Assuming out is a list with elements fnFs and fnRs containing the paths to trimmed files
```

##Quality Control after Trimming

```{r pressure, echo=FALSE}
plotQualityProfile(fnFs2[1:2])  # # Trimmed forward reads
plotQualityProfile(fnRs2[1:2])  # Trimmed Reverse reads
```

```{r pressure, echo=FALSE}
#Learn the Error Rates
errF <- learnErrors(filtFs, multithread=TRUE) 
errR <- learnErrors(filtRs, multithread=TRUE)
plotErrors(errF, nominalQ=TRUE)

#Sample Inference
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
dadaFs[[1]]

#Merge paired reads
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

#Construct sequence table
#We can now construct an amplicon sequence variant table (ASV) table
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
#The sequence table is a matrix with rows corresponding to (and named by) the samples, and columns corresponding to (and named by) the sequence variants.

#Remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
#sum(seqtab.nochim)/sum(seqtab)
```

##filtering the ASV table removing water related MOs

```{r pressure, echo=FALSE}
# Identify rows corresponding to H2O samples
h2o_samples <- c("H2O-H1", "H2O-H2", "H2O-H3")

# Step 1: Find the rows corresponding to H2O samples
h2o_rows <- rownames(seqtab.nochim)[grepl("H2O-H[1-3]", rownames(seqtab.nochim))]
h2o_rows
# Step 2: Find columns where values are not equal to 0 for the identified rows
non_zero_columns <- colnames(seqtab.nochim)[apply(seqtab.nochim[h2o_rows, ], 2, function(x) any(x != 0))]

# Step 3: Filter out these columns for all rows
filtered_seqtab <- seqtab.nochim[, -which(colnames(seqtab.nochim) %in% non_zero_columns)]
# Omit rows corresponding to H2O samples
filtered_seqtab <- filtered_seqtab[!grepl("H2O-H[1-3]", rownames(filtered_seqtab)), ]



# Step 4: Save the filtered matrix table to a new file
write.table(filtered_seqtab, file = "05_processed_data/filt_seqtab_nochim.csv", sep = ",", quote = FALSE)


```

```{r pressure, echo=FALSE}
#Track reads through the pipeline
#As a final check of our progress, we’ll look at the number of reads that made it through each step in the pipeline:
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

#Assign taxonomy
taxa <- assignTaxonomy(filtered_seqtab, "~/fallstudie/silva_nr99_v138.1_train_set.fa.gz", multithread=TRUE)
taxa <- addSpecies(taxa, "~/fallstudie/silva_species_assignment_v138.1.fa.gz")
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

#Saving the Result

```{r pressure, echo=FALSE}
# Save the results
#write.table(taxa.print, file="05_processed_data/taxa_print.txt", sep="\t")
write.table(filtered_seqtab, file="05_processed_data/filtered_seqtab.txt", sep="\t")
#write.table(taxa, file="05_processed_data/taxonomy.txt", sep="\t")
```

#Phyloseq

```{r pressure, echo=FALSE}
library(phyloseq)
library(readxl)

# add EOTRH as dataframe

metadata <- read_excel("EOTRH-MetadatenProben_studvers_2023.xlsx", 
    skip = 1)

# Set SeqPos as row names
#rownames(taxa) 
#Remove the SeqPos column from the data frame if needed
#transpose
filtered_seqtab <- t(filtered_seqtab)

#check if the sample data is correct
#colnames(filtered_seqtab) %in% metadata$SeqPos
#rownames(filtered_seqtab) %in% rownames(taxa)


# it should be true
all(rownames(filtered_seqtab) %in% rownames(taxa))
#cleaning the data 
nrow(metadata)
metadata <- metadata [1:15,]
metadata$SeqPos <- gsub("-", "_", metadata$SeqPos, fixed=TRUE)
colnames(filtered_seqtab) <- gsub("-", "_", colnames(filtered_seqtab), fixed=TRUE)
metadata <- data.frame(metadata)
rownames(metadata) <- metadata$SeqPos

# Create the phyloseq object
physeq <- phyloseq(otu_table(filtered_seqtab, taxa_are_rows = TRUE),
                   sample_data(metadata),
                   tax_table(taxa))


#Check if phyloseq object was created correctly
physeq
nsamples(physeq) # number of samples
ntaxa(physeq) # Number of taxa (ASV)
sample_variables(physeq) # Number of sample variables
rank_names(physeq) # ranking level

# Print summary of the phyloseq object
print(physeq)
#save result
write.csv(otu_table(physeq), file = "05_processed_data/physeq_table.csv")
#short names for our ASVs
#physeq <- merge_phyloseq(physeq, dna)
#taxa_names(physeq_table) <- short_names
Sequencedepth <- as.matrix(sample_sums(physeq))
physeq = prune_samples(sample_sums(physeq)>=10,
                               physeq)
Sequencedepth <- as.matrix(sample_sums(physeq))


#mean sequenced reads per sample after filtering
mean(as.matrix(sample_sums(physeq)))

#mean number of sequenced reads per ASV after filtering
mean(as.matrix(taxa_sums(physeq)))
nsamples(physeq) # number of samples

```

### Analysis

```{r pressure, echo=FALSE}
#load libraries
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")

# Define new phyloseq object with relative abundances (nomralization)
physeq_rel = transform_sample_counts(physeq, function(x) x / sum(x)*100)


#######extract files with taxon name
#Extract abundance matrix from the phyloseq object with the taxon name as the row
ASV1 = as.data.frame(otu_table(physeq))

#create ASV2 matrix that contains the ASV names as row names and the corresponding taxonomy
ASV2 = as.data.frame(tax_table(physeq))

#combine the two matrices together by appending the rows together 
ASV3 = cbind(ASV1, ASV2)
ASV3[ASV3==0]<- NA # change zero values to NA

#make the rownames into the first column of the matrix
df <- data.frame(ASVs = row.names(ASV3), ASV3)
rownames(df) = NULL
str(ASV3)

# save the table
library (writexl)
write_xlsx(df,"05_processed_data/phyloseq_ASVtable_with_ASVs.xlsx") 


# Pivot (Genus)
#install.packages("tidyverse")
library (tidyverse)
library(dplyr)
ASV4 <-
  ASV3 %>% 
  select(-c(Kingdom, Phylum, Class, Order,Family)) %>% 
  pivot_longer(!c(Genus, Species),
               names_to="Sample", values_to="asv_count") %>% 
  as.data.frame()


head(ASV4)


ASV5 <- aggregate(asv_count~Genus+Sample, data=ASV4, FUN=mean)
head(ASV5)

ASVF<- pivot_wider(data=ASV5, names_from=Genus, values_from=asv_count)



library(funrar)
ASVF1 <- ASVF[,-1] # remove the sample names for matrix
ASVF1 <- as.matrix(ASVF1) # change it into matrix 
rel_ASVF1 <- make_relative(ASVF1)  #calculate the relative abundance
rel_ASVF1 <- rel_ASVF1*100 # multiply by 100
rel_ASVF1 <- cbind(ASVF[1],rel_ASVF1) # add sample names to the dataframe


# save the table
 library (writexl)
 write_xlsx(rel_ASVF1,"05_processed_data/phyloseq_ASVtable_with_Genus.xlsx") 
 
 
 
 # Pivot (Species)
 library (tidyverse)
 ASV4 <-
   ASV3 %>% 
   select(-c(Kingdom, Phylum, Class, Order, Family)) %>% 
   pivot_longer(!c(Genus, Species),
                names_to="Sample", values_to="asv_count") %>% 
   as.data.frame()
 
 
 head(ASV4)
 
ASV5 <- aggregate(asv_count~Species+Sample, data=ASV4, FUN=mean)
 head(ASV5)
 
 ASVF<- pivot_wider(data=ASV5, names_from=Species, values_from=asv_count)
 
 
 library (writexl)
 write_xlsx(ASVF,"05_processed_data/phyloseq_ASVtable_with_Species.xlsx") 
 
 
# Filter rows where Genus is equal to "Tannerella"
tannerella_samples <- ASV4 %>%
  filter(Genus == "Tannerella")

# Display the resulting data frame
print(tannerella_samples)
# Filter rows where asv_count is not NA
valid_samples <- tannerella_samples[!is.na(tannerella_samples$asv_count), ]

# Create a bar plot
ggplot(valid_samples, aes(x = Sample, y = asv_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "ASV Counts for Tannerella Samples",
       x = "Sample",
       y = "ASV Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("06_result/Tannerella Samples.png", width = 10, height = 6)


# Check for unknown microorganisms in the "Genus" column
unknown_microorganisms <- ASV4$Genus %in% c(NA, "unclassified", "unannotated", "unknown")

# Count the occurrences of unknown microorganisms
num_unknown_microorganisms <- sum(unknown_microorganisms)

# Display the result
cat("Number of unknown microorganisms in horses:", num_unknown_microorganisms, "\n")


```

### Multivariate analysis

```{r pressure, echo=FALSE}

#Show ASV in NMDS (different ranks)
GP.ord <- ordinate(physeq_rel, "NMDS", "jaccard")
p1 = plot_ordination(physeq_rel, GP.ord, type="taxa", color="Family", title="taxa")
print(p1)
ggsave("06_result/NMDS_ASV.jpg", height=5)

# show NMDS in rank groups
p1 + facet_wrap(~Family, 3)
ggsave("06_result/NMDS_ASV_Ranking.jpg", width=9, height=5)







# Samples in NMDS
p2 = plot_ordination(physeq_rel, GP.ord, color="diseasestate") + 
  ggtitle("diseasestate - NMDS: Jaccard") +
  scale_color_manual(values=c("#39F3FC", "#518F42", "#FEDA6A", "#EE5E00",
                             "#A7FF92", "#0A45E2", "#F1AAA4", "#CB3D0C",
                             "#D7D5D5", "#7D7878", "#090A0B", "#CCB8A3",
                             "#A88661", "#5C4833"
  )) +
  theme(legend.title = element_blank())
p2
ggsave("06_result/NMDS_plot_disease.jpg", width=10, height=5)


p2 + geom_polygon(aes(fill=diseasestate)) + geom_point(size=3) + 
  ggtitle("diseasestate - NMDS: Jaccard") +
  scale_fill_manual(values=c("#39F3FC", "#518F42", "#FEDA6A", "#EE5E00",
                             "#A7FF92", "#0A45E2", "#F1AAA4", "#CB3D0C",
                             "#D7D5D5", "#7D7878", "#090A0B", "#CCB8A3",
                             "#A88661", "#5C4833"
  )) +
  theme(legend.title = element_blank())

ggsave("06_result//NMDS_polygon_diseasestate.jpg", width=10, height=5)


library(phyloseq)
install.packages("phyloseq")

##PCoA
# Calculate distances
DistBC = distance(physeq_rel, method = "jaccard")


#Perform "PCoA" Ordination
ordBC = ordinate(physeq_rel, method = "PCoA", distance = DistBC)

#Before exploring the PCoA plot, let's explore a property of each ordination itself. The eigenvalues associated with each axis indicate the relative proportion of total variability within the distance matrix that is represented in that axis. The axes are already ordered from most variability to least, and so a plot of these values in order of each axis gives a helpful impression as to which, and how many, axes we should bother plotting and interpreting.
plot_scree(ordBC, "Scree Plot: Jaccard")

#Finally, let's plot a PCoA the distance-and-ordination using the plot_ordination() function in phyloseq.
PcoA <- plot_ordination(physeq_rel, ordBC, color = "diseasestate") + 
  geom_point(size=3)+
  ggtitle("diseasestate - PCoA: Jaccard") +
scale_color_manual(values=c("#39F3FC", "#518F42", "#FEDA6A", "#EE5E00",
                                     "#A7FF92", "#0A45E2", "#F1AAA4", "#CB3D0C",
                                     "#D7D5D5", "#7D7878", "#090A0B", "#CCB8A3",
                                     "#A88661", "#5C4833"
)) 
PcoA + theme(legend.title = element_blank())

ggsave("06_result/PCoA_plot_jaccard_disease.jpg", width=8, height=5)



##### PcoA Polygon
PcoA + geom_polygon(aes(fill=diseasestate)) +
  geom_point(size=3) +
  ggtitle("diseasestate - PCoA: Jaccard") +
  scale_fill_manual(values=c("#39F3FC", "#518F42", "#FEDA6A", "#EE5E00",
                             "#A7FF92", "#0A45E2", "#F1AAA4", "#CB3D0C",
                             "#D7D5D5", "#7D7878", "#090A0B", "#CCB8A3",
                             "#A88661", "#5C4833")) +
  theme(legend.title = element_blank())
ggsave("06_result/PCoA_polygon_jaccard_disease.jpg", width=8, height=5)





```

###CA method

```{r pressure, echo=FALSE}
DistCA = distance(physeq_rel, method = "jaccard")

#Perform "CA" Ordination
ordCA = ordinate(physeq_rel, method = "CCA", distance = DistCA)

#Before exploring the CA plot, let's explore a property of each ordination itself. The eigenvalues associated with each axis indicate the relative proportion of total variability within the distance matrix that is represented in that axis. The axes are already ordered from most variability to least, and so a plot of these values in order of each axis gives a helpful impression as to which, and how many, axes we should bother plotting and interpreting.
plot_scree(ordCA, "Scree Plot: Jaccard")

#Finally, let's plot a CA the distance-and-ordination using the plot_ordination() function in phyloseq.
CA <- plot_ordination(physeq_rel, ordCA, color = "Type") + geom_point(size=3)+
  ggtitle("Type - CA: Jaccard") +
  scale_color_manual(values=c("#39F3FC", "#518F42", "#FEDA6A", "#EE5E00",
                              "#A7FF92", "#0A45E2", "#F1AAA4", "#CB3D0C",
                              "#D7D5D5", "#7D7878", "#090A0B", "#CCB8A3",
                              "#A88661", "#5C4833"
  )) 
CA +  theme(legend.title = element_blank())
ggsave("06_result/CA_plot_jaccard_type.jpg", width=8, height=5)




```

## 

For that reason we'll store the DNA sequences of our ASVs in the refseq slot of the phyloseq object, and then rename our taxa to a short string. That way, the short new taxa names will appear in tables and plots, and we can still recover the DNA sequences corresponding to each ASV as needed with refseq(ps).

```{r pressure, echo=FALSE}

dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq, dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq
tax_table(physeq)

plot_richness_diseasestate <- plot_richness(physeq, measures=c("Shannon", "Simpson"), color="diseasestate")
ggsave("06_result/plot_richness_diseasestate.png", plot = plot_richness_diseasestate)

plot_richness_Type <- plot_richness(physeq, measures=c("Shannon", "Simpson"), color="Type")
ggsave("06_result/plot_richness_Type.png", plot = plot_richness_Type)

# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")

Bray_NMDS_diseasestate <-plot_ordination(ps.prop, ord.nmds.bray, color="diseasestate", title="Bray NMDS")
ggsave("06_result/Bray_NMDS_diseasestate.png", plot = Bray_NMDS_diseasestate)

Bray_NMDS_type <-plot_ordination(ps.prop, ord.nmds.bray, color="Type", title="Bray NMDS")
ggsave("06_result/Bray_NMDS_type.png", plot = Bray_NMDS_type )

top20 <- names(sort(taxa_sums(physeq), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(physeq, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
Plotbar_diseasestate <-plot_bar(ps.top20, , fill="Family" ) + facet_wrap(~diseasestate, scales="free_x")
ggsave("06_result/Plotbar_diseasestate.png", plot = Plotbar_diseasestate)

Plotbar_type <-plot_bar(ps.top20, , fill="Family" ) + facet_wrap(~Type, scales="free_x")
ggsave("06_result/Plotbar_type.png", plot = Plotbar_type)
```
